---
title: "EC242 Group Project 1"
author:
  - "Group Name: ACJ_MSU"
  - "Group Members: Anish Poudel, Joseph Pelkey, Cooper Scheble"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
---

```{r LibraryChunk, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.align = 'center')
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## **Part 1** 

### **Rat Sightings Analysis**

### **Data Processing & Cleaning**

The dataset was cleaned by standardizing date formats, extracting relevant time-based variables, and filtering out missing borough data to ensure accurate analysis.

```{r}
# Packages that we will be using
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(sf)
library(broom)
library(effects)

# Data Read
rats_raw <- read_csv("Rat_sightings.csv", na = c("", "NA", "N/A"))
rats_clean <- rats_raw %>%
  rename(created_date = `Created Date`,
         location_type = `Location Type`,
         borough = Borough) %>%
  mutate(created_date = mdy_hms(created_date)) %>%
  mutate(sighting_year = year(created_date),
         sighting_month = month(created_date),
         sighting_day = day(created_date),
         sighting_weekday = wday(created_date, label = TRUE, abbr = FALSE)) %>%
  filter(borough != "Unspecified")
```


```{r}
library(knitr)
library(kableExtra)
rats_summary <- rats_clean %>%
  group_by(borough, sighting_year) %>%
  summarise(sightings = n(), .groups = "drop")
rats_wide <- rats_summary %>%
  pivot_wider(
    id_cols = borough,
    names_from = sighting_year,
    values_from = sightings
  )
rats_wide %>%
  kable(caption = "Rat Sightings by Borough and Year", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Year" = ncol(rats_wide) - 1))
```

Our analysis showed that:

- Brooklyn consistently had the highest sightings, maximum in 2017.

- The Bronx and Manhattan showed increasing trends from 2014 to 2017.

- Queens and Staten Island had significantly lower complaints.

- Sightings dropped in 2020, which might be due to COVID-19 pandemic changes.

These trends show that certain boroughs require more targeted pest control strategies.

\newpage

```{r}
rats_summary <- rats_clean %>%
  group_by(borough, sighting_year) %>%
  summarise(sightings = n(), .groups = "drop")
# Heatmap
ggplot(rats_summary, aes(x = sighting_year, y = borough, fill = sightings)) +
  geom_tile(color = "white") +  
  geom_text(aes(label = sightings), color = "black", size = 4) +  
  scale_fill_gradient(low = "white", high = "darkred") + 
  theme_minimal() +
  labs(title = "Heatmap of Rat Sightings in NYC by Borough and Year",
       x = "Year",
       y = "Borough",
       fill = "Sightings") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9))
```

The heatmap clearly shows the following:

- Brooklyn and Manhattan have the most rat sightings.

- There were major increases in rat reports during 2015-2017 throughout several boroughs.

- Queens and Staten Island have fewer rat sightings overall.

Based on this, we can say that pest control teams should concentrate their efforts in the boroughs with the highest number of rats, mainly Brooklyn and Manhattan.

\newpage

```{r}
# Line plot
ggplot(rats_summary, aes(x = sighting_year, y = sightings, color = borough)) +
  geom_line(linewidth = 1) +
  geom_point() +
  theme_minimal() +
  labs(title = "Trend of Rat Sightings in NYC (2010-2020)",
       x = "Year",
       y = "Number of Sightings",
       color = "Borough")
```

```{r}
Zip_Code_Raw <- read_csv("Modified_Zip_Code_Tabulation_Areas__MODZCTA_.csv")
income <- read_csv("income_data.csv") %>%
  rename(ZIP = geoid, Median_Income = B19013001)
pop <- read_csv("populations.csv") %>%
  rename(ZIP = geoid, Population = B01003001)
merged <- income %>%
  inner_join(pop, by = "ZIP") %>%
   filter(!is.na(Median_Income), !is.na(Population)) %>%
  select(-"B01003001, Error", -"B19013001, Error")


merged <- dplyr::select(merged, -ZIP, -name.y) %>%
  rename(ZIP = name.x)
merged <- merged %>%
  filter(Population > 10000)



names(Zip_Code_Raw)[1] <- ("Zip")

```


```{r}
merged <- merged %>%
  filter(!ZIP %in% c("New York, NY"))
```


```{r}
print(merged)
```


```{r}
rats_zip_time <- rats_clean %>%
  mutate(
    created = as.Date(created_date),
    Year = year(created),
    ZIP = stringr::str_extract(`Incident Zip`, "\\d{5}")
  ) %>%
  filter(!is.na(ZIP), ZIP != "00000")
```


```{r}
rat_yearly <- rats_clean %>%
  mutate(
    created = as.Date(created_date),
    Year = year(created),
    ZIP = stringr::str_extract(`Incident Zip`, "\\d{5}")
  ) %>%
  filter(!is.na(ZIP), ZIP != "00000") %>%
  group_by(ZIP, Year, borough) %>%   # <-- include borough here
  summarise(Sightings = n(), .groups = "drop")

```

```{r}
full_dataset <- rat_yearly %>%
  group_by(ZIP) %>%
  summarise(
    Avg_Sightings_per_Year = mean(Sightings),
    borough = first(borough),              # keep borough
    .groups = "drop"
  ) %>%
  inner_join(merged, by = "ZIP") %>%
  mutate(
    Sightings_per_1000_per_Year = Avg_Sightings_per_Year / Population * 1000
  )
```


```{r}
zips_sf <- st_as_sf(Zip_Code_Raw, wkt = "the_geom", crs = 4326)
zips_sf$Zip <- as.character(zips_sf$Zip)
zips_sf <- st_make_valid(zips_sf)
ggplot(zips_sf) +
  geom_sf(fill = "grey95", color = "grey30", size = 0.2) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 12) +
  labs(title = "NYC ZIP Code Boundaries", caption = "Source: NYC Open Data")
```

```{r}
zips_sf <- zips_sf %>%
#rename(ZIP = Zip) %>%             # rename if needed
  mutate(ZIP = as.character(ZIP))

full_dataset <- full_dataset %>%
  mutate(ZIP = as.character(ZIP))
```

```{r}
full_dataset_sf <- full_dataset %>%
  left_join(
    zips_sf %>% select(ZIP, the_geom),
    by = "ZIP"
  ) %>%
  st_as_sf()

```


```{r}
full_dataset_sf <- st_transform(full_dataset_sf, 2263)


full_dataset_sf <- full_dataset_sf %>%
  mutate(
    area_sqft = as.numeric(st_area(the_geom)),         # square feet
    area_sqmi = area_sqft / (5280^2),                  # convert to square miles
    area_sqkm = area_sqft / 10.7639 / 1e6,             # convert to square km
    pop_density_sqmi = Population / area_sqmi,         # people per square mile
    pop_density_sqkm = Population / area_sqkm          # people per square km
  )



```


```{r}
full_dataset_sf <- full_dataset_sf %>%
  filter(!ZIP %in% c(11001, 11040, 11249))
```


```{r}



m_ols <- lm(Sightings_per_1000_per_Year ~ Median_Income + pop_density_sqkm, 
            data = full_dataset_sf)

summary(m_ols)
```


```{r}
zips_keep <- unique(full_dataset_sf$ZIP)
rats_filtered <- rat_yearly %>%
  filter(ZIP %in% zips_keep)

```

```{r}
zip_density <- full_dataset_sf %>%
  st_drop_geometry() %>%
  select(ZIP, Sightings_per_1000_per_Year, pop_density_sqkm, Median_Income)

rats_with_density <- rats_filtered %>%
  left_join(zip_density, by = "ZIP")

head(rats_with_density)
```

```{r}
m_year_borough <- lm(
  Sightings_per_1000_per_Year ~ Median_Income + pop_density_sqkm +
    factor(Year) + factor(borough),
  data = rats_with_density
)



tidy(m_year_borough, conf.int = TRUE) %>%
  filter(term %in% c("Median_Income", "pop_density_sqkm", 
                     "factor(borough)BROOKLYN", 
                     "factor(borough)MANHATTAN", 
                     "factor(borough)QUEENS", 
                     "factor(borough)STATEN ISLAND")) %>%
  ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  coord_flip() +
  labs(x = "", y = "Coefficient estimate",
       title = "Effect of Income, Density, and Borough on Rat Sightings")
```


```{r}
library(car)
avPlots(m_year_borough, terms = ~ pop_density_sqkm + Median_Income)
```

