---
title: "Urban Density And Rat Sightings: Evidence from New York City Zip Codes"
author: "John Scheble"
project:
    type: website 
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  message: false
  warning: false
  echo: false  
---

**This report explores NYC rat sightings by ZIP, linking complaints to population, income, and area to get density-adjusted rates. Goal is to work with sightings per 1000 residents and population density to comapre neighborhoods fairly. Data is from NYC 3-1-1 Rat Complaints.**



# **Data Processing & Cleaning**

The dataset was cleaned by standardizing date formats, extracting relevant time-based variables, and filtering out missing borough data to ensure accurate analysis.

```{r setup, echo=FALSE, warning=FALSE}
# Packages that we will be using
library(tidyverse); library(lubridate); library(dplyr); library(ggplot2); library(sf); library(broom); library(effects); library(ggokabeito); library(ggtext); library(ggrepel); library(ggpubr); library(scales); library(knitr); library(kableExtra); library(viridis); library(classInt); library(performance); library(see); library(gtsummary); library(modelsummary)
```

```{r, echo=FALSE, message=FALSE}
# Data Read and Renaming
rats_raw <- read_csv("Rat_sightings.csv", na = c("", "NA", "N/A"))
rats_clean <- rats_raw %>%
  rename(created_date = `Created Date`,
         location_type = `Location Type`,
         borough = Borough) %>%
  mutate(created_date = mdy_hms(created_date)) %>%
  mutate(sighting_year = year(created_date),
         sighting_month = month(created_date),
         sighting_day = day(created_date),
         sighting_weekday = wday(created_date, label = TRUE, abbr = FALSE)) %>%
  filter(borough != "Unspecified")
```

# Tile Heat Map by Borough by Year

```{r}
#Summarize the sightings by borough
rats_summary <- rats_clean %>%
  group_by(borough, sighting_year) %>%
  summarise(sightings = n(), .groups = "drop")
# Heatmap
ggplot(rats_summary, aes(x = sighting_year, y = borough, fill = sightings)) +
  geom_tile(color = "white") +  
  geom_text(aes(label = sightings), color = "black", size = 4) +  
  scale_fill_gradient(low = "white", high = "darkred") + 
  theme_minimal() +
  labs(title = "Heatmap of Rat Sightings in NYC by Borough and Year",
       x = "Year",
       y = "Borough",
       fill = "Sightings") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 9))
```

The heatmap shows the following:

-   Brooklyn and Manhattan have the most rat sightings.

-   There were major increases in rat reports during 2015-2017 throughout several boroughs.

-   Queens and Staten Island have fewer rat sightings overall.

Based on this, we can say that pest control teams should concentrate their efforts in the boroughs with the highest number of rats, mainly Brooklyn and Manhattan.

# Rat Sightings Per Borough Over Time

```{r}
# Line plot
ggplot(rats_summary, aes(x = sighting_year, y = sightings, color = borough)) +
  geom_line(linewidth = 1) +
  geom_point() +
  theme_minimal() +
  labs(title = "Trend of Rat Sightings in NYC (2010-2020)",
       x = "Year",
       y = "Number of Sightings",
       color = "Borough")
```

The above graph illustrates the large and steady quantity difference in rat sightings between boroughs over time. We see that Brooklyn has consistently had the largest number of rat sightings for each year recorded, followed by Manhattan, Bronx, Queens and Staten Island Respectively. Over time, there hasnt been much change in this order, with queens slightly overtaking the bronx in 2020. Also, between 2013 and 2018 there seemed to be an large influx of rat sightings across the board (excluding Staten Island which remained relatively unchanged.)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Adding extra zip code df, median family income by zip, and zip populations.
Zip_Code_Raw <- read_csv("Modified_Zip_Code_Tabulation_Areas__MODZCTA_.csv")
income <- read_csv("income_data.csv") %>%
  rename(ZIP = geoid, Median_Income = B19013001)
pop <- read_csv("populations.csv") %>%
  rename(ZIP = geoid, Population = B01003001)
merged <- income %>%
  inner_join(pop, by = "ZIP") %>%
   filter(!is.na(Median_Income), !is.na(Population)) %>%
  select(-"B01003001, Error", -"B19013001, Error")

#Merged
merged <- dplyr::select(merged, -ZIP, -name.y) %>%
  rename(ZIP = name.x)
merged <- merged %>%
  filter(Population > 10000)



names(Zip_Code_Raw)[1] <- ("Zip")

```

```{r}
#Remove extra row
merged <- merged %>%
  filter(!ZIP %in% c("New York, NY"))
```

```{r}
#Reclean First Rat Dataset
rats_zip_time <- rats_clean %>%
  mutate(
    created = as.Date(created_date),
    Year = year(created),
    ZIP = stringr::str_extract(`Incident Zip`, "\\d{5}")
  ) %>%
  filter(!is.na(ZIP), ZIP != "00000")
```

```{r}
#Sighting per year per bourough dataset
rat_yearly <- rats_clean %>%
  mutate(
    created = as.Date(created_date),
    Year = year(created),
    ZIP = stringr::str_extract(`Incident Zip`, "\\d{5}")
  ) %>%
  filter(!is.na(ZIP), ZIP != "00000") %>%
  group_by(ZIP, Year, borough) %>%   # <-- include borough here
  summarise(Sightings = n(), .groups = "drop")

```

```{r}
#Combined Dataset
full_dataset <- rat_yearly %>%
  group_by(ZIP) %>%
  summarise(
    Avg_Sightings_per_Year = mean(Sightings),
    borough = first(borough),              # keep borough
    .groups = "drop"
  ) %>%
  inner_join(merged, by = "ZIP") %>%
  mutate(
    Sightings_per_1000_per_Year = Avg_Sightings_per_Year / Population * 1000
  )
```

```{r}
#loading in zip shapefile
zips_sf <- st_as_sf(Zip_Code_Raw, wkt = "the_geom", crs = 4326)
zips_sf$Zip <- as.character(zips_sf$Zip)
zips_sf <- st_make_valid(zips_sf)
```

```{r}
#Adding geometry to zips
zips_sf <- zips_sf %>%
rename(ZIP = Zip) %>%             
 mutate(ZIP = as.character(ZIP))

full_dataset <- full_dataset %>%
  mutate(ZIP = as.character(ZIP))
```

```{r}
#Adding geometry to full dataset
full_dataset_sf <- full_dataset %>%
  left_join(
    zips_sf %>% select(ZIP, the_geom),
    by = "ZIP"
  ) %>%
  st_as_sf()

full_dataset_sf <- full_dataset_sf %>%
  filter(!ZIP %in% c(11001, 11040, 11249))
```

```{r}
#Calculating area of zip codes in sqft, miles, sqkm, and then pop density
full_dataset_sf <- st_transform(full_dataset_sf, 2263)


full_dataset_sf <- full_dataset_sf %>%
  mutate(
    area_sqft = as.numeric(st_area(the_geom)),         # square feet
    area_sqmi = area_sqft / (5280^2),                 #Square Miles
    area_sqkm = area_sqft / 10.7639 / 1e6,             #square km
    pop_density_sqmi = Population / area_sqmi,         #people per square mile
    pop_density_sqkm = Population / area_sqkm          #people per square km
  )
```


# Choropleth Map of Sightings

```{r}
#Create heatmap with zip plotting
if (!inherits(full_dataset_sf, "sf")) {
  full_dataset_sf <- st_as_sf(full_dataset_sf, wkt = "the_geom", crs = 4326)
}


full_dataset_sf_plot <- st_transform(full_dataset_sf, 3857)


metric <- "Sightings_per_1000_per_Year"

p_cont <- ggplot(full_dataset_sf_plot) +
  geom_sf(aes(fill = .data[[metric]]), color = "white", linewidth = 0.2) +
  scale_fill_viridis(
    option = "C", direction = 1,
    na.value = "grey90",
    name = label_wrap(30)("Average # of Rats Reported Every Year, Per 1000 People")
  ) +
  labs(
    title = "Rat Sightings by ZIP Code",
    subtitle = "Heatmap (continuous scale)",
    caption = "Source: NYC 311 Rat Sightings; Author’s calculations"
  ) +
  theme_void(base_size = 12) +
  theme(
    legend.title = element_text(size = 12, face = "bold"),  # bigger title
    legend.text  = element_text(size = 11),                 # bigger labels
    legend.key.width = unit(0.4, "cm"),                     # shrink key width
    legend.key.height = unit(0.8, "cm")                     # taller key
  )
```



```{r}
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Rat sightings per 1,000 residents by ZIP"
#| out-width: 100%
p_cont   # <- simply print the ggplot object
```

The above heat map illustrates exactly where rat sightings are in the 5 bouroughs. It is calculated by averaging total rat calls by the 10 years the data was collected, then adjusting per 1000 people in order to standardize for population.

# Regression Model Results

```{r}

dat <- full_dataset_sf |>
  st_drop_geometry() |>
  mutate(
    pop_density_k = pop_density_sqkm / 1000,     # per 1,000 people/km²
    Median_Income = Median_Income,      
    borough = fct_relevel(borough, "BRONX")      # pick your reference explicitly
  )

m0  <- lm(Sightings_per_1000_per_Year ~ Median_Income + pop_density_k, data = dat)
mFE <- lm(Sightings_per_1000_per_Year ~ Median_Income + pop_density_k + borough, data = dat)

modelsummary(
  list("Baseline" = m0, "Borough FE" = mFE),
  vcov = "HC3",                      # robust SEs
  statistic = "({std.error})",
  stars = TRUE,
  coef_map = c(
    "(Intercept)"    = "Intercept",
    "Median_Income"= "Median income (per $1k)",
    "pop_density_k"  = "Pop. density (per 1k/km²)"
  ),
  gof_map = c("nobs","r.squared","adj.r.squared","AIC","BIC","rmse"),
  output = "markdown"
)
```

The results from regression shows that population density is positively associated with complaint rates. In the baseline model, a 1,000 person/km\^2 increase in density predicts +0.040 additional sightings per 1,000 residents per year. When you add Boroughs as fixed effects, this effect by population density decreases to 0.029. Median income is indistinguishable from zero across models, and is not statistically signifigant. Adding borough indicators imrpoved the fit from r: 0.197 to 0.301. Estimates are descriptive.

```{r}
#Filter and create new dataset with key columns
zips_keep <- unique(full_dataset_sf$ZIP)
rats_filtered <- rat_yearly %>%
  filter(ZIP %in% zips_keep)

zip_density <- full_dataset_sf %>%
  st_drop_geometry() %>%
  select(ZIP, Sightings_per_1000_per_Year, pop_density_sqkm, Median_Income)

rats_with_density <- rats_filtered %>%
  left_join(zip_density, by = "ZIP")

```

# Correlation Scatterplot

```{r}
#Correlation scatterplot
ggscatter(
  rats_with_density,
  x = "pop_density_sqkm",
  y = "Sightings_per_1000_per_Year",
  add = "reg.line",
  conf.int = TRUE,
  color = "#0072B2",  
  add.params = list(color = "#D55E00", linewidth = 1.1),
  xlab = "Population density (people per km²)",
  ylab = "Rat sightings per 1,000 residents",
  title = "Correlation between Density and Rat Sightings"
) +
  stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top") +
  theme_pubr(base_size = 12) +
  theme(legend.position = "none")
```

Denser ZIP codes report more rat sightings per capita. The relationship is moderate and positive (R = 0.45), consistent with the fitted line and confidence band. Dispersion widens at higher densities, so effects are not uniform across neighborhoods.

# Correlation of Sightings with Density vs Income

```{r}
#1st plot density vs sightings
p_density <- ggscatter(
  rats_with_density,
  x = "pop_density_sqkm",
  y = "Sightings_per_1000_per_Year",
  add = "reg.line",
  conf.int = TRUE,
  xlab = "Population density (people per km²)",
  ylab = "Rat sightings per 1,000 residents",
  title = "Density vs Sightings"
) +
  stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top") +
  scale_x_continuous(labels = label_comma()) +
  theme_pubr(base_size = 12) +
  theme(legend.position = "none")



#2nd plot Income vs Sightings
p_income <- ggscatter(
  rats_with_density,
  x = "Median_Income",
  y = "Sightings_per_1000_per_Year",
  add = "reg.line",
  conf.int = TRUE,
  xlab = "Median household income (USD)",
  ylab = "Rat sightings per 1,000 residents",
  title = "Income vs Sightings"
) +
  stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top") +
  scale_x_continuous(labels = label_dollar(accuracy = 1)) +
  theme_pubr(base_size = 12) +
  theme(legend.position = "none")


# Sie by SIde for dramatic effect
paired <- ggarrange(
  p_density, p_income,
  ncol = 2, nrow = 1,
  labels = c("A", "B"),
  align = "hv"
)

annotate_figure(
  paired,
  top = text_grob("Correlation of Rat Sightings with Density and Income",
                  face = "bold", size = 14),
  bottom = text_grob("Points = ZIPs; line = OLS with 95% CI", size = 10)
)
```

Panel A shows a clear, positive association: denser ZIPs tend to report more rat sightings per capita. The spread widens at higher densities (fan shape), so variance is larger in dense areas—reason to use robust SEs in regressions. Panel B is essentially flat: income is not correlated with sightings in these data (R ≈ 0, p ≈ 0.99). Any relationship between income and complaints is either negligible or swamped by other factors like density, building stock, or reporting behavior.

# Mapped Excess Rat Sightngs Relative to Density
```{r}
#| label: residual-map
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

library(colorspace)


df_mod <- st_drop_geometry(full_dataset_sf)
mod <- lm(Sightings_per_1000_per_Year ~ log1p(pop_density_sqkm) + borough, data = df_mod)


df_mod$pred   <- predict(mod, newdata = df_mod)
df_mod$resid  <- df_mod$Sightings_per_1000_per_Year - df_mod$pred
df_mod$resid_z <- (df_mod$resid - mean(df_mod$resid, na.rm = TRUE)) / sd(df_mod$resid, na.rm = TRUE)

# Join back to geometry
map_resid <- full_dataset_sf |> select(ZIP, geometry = the_geom) |>
  left_join(df_mod |> select(ZIP, resid_z), by = "ZIP")

ggplot(map_resid) +
  geom_sf(aes(fill = resid_z), color = "white", linewidth = 0.25) +
scale_fill_continuous_divergingx(
    palette = "PuOr",   
    mid = 0,                  
    limits = c(-3, 3),
    breaks = c(-2, -1, 0, 1, 2),
    oob = scales::squish,
    na.value = "grey95",
    name = "Excess sightings\n(rel. to density)\nZ-score"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text  = element_text(size = 10)
  ) +
  guides(fill = guide_colorbar(title.position = "top",
                               barheight = unit(4, "cm"),
                               barwidth  = unit(0.5, "cm")))
```

This map highlights where sightings are unusually high or low after accounting for density. Dark purples represent ZIPs that have larger than expected complaint rates given how dense they are, while oranges mark places that report less than expected. Clustering suggests localized conditions like building stock, sanitation practices, enforcement, or reporting behavior influence sightings instead of density alone.
